---
title: "Day 2 Lectures"
author: "AJ Smit"
format:
  html: default
  typst: default
  docx: default
---

# This is a heading

This is the same comment that I hash out inside the code block: load the tidyverse library

```{r}
# load the tidyverse library
library(tidyverse)
```

```{r}
mean(1, 2, 3)
```

## Numeric data

```{r}
# Create a vector of the number of pets owned by each household
num_rats <- c(0, 1, 2, 2, 3, 1, 4, 0, 2, 1, 2, 2, 0, 3, 2, 1, 1, 4, 2, 0)
num_rats
```

```{r}
class(num_rats)
```

```{r}
num_rats_int <- as.integer(num_rats) # coercion
num_rats_int
```

```{r}
class(num_rats_int)
```

```{r}
pies <- pi * seq(1:5)
pies
```

```{r}
class(pies)
```

```{r}
as.integer(pies)
```

```{r}
round(pies, 2)
```

The `as.integer()` function does not care about rouning, for example:

```{r}
as.integer(round(pies, 2))
```

```{r}
set.seed(13)
# Generate a vector of 50 normally distributed temperature values
temp_data <- round(rnorm(n = 50, mean = 15, sd = 3), 2)
# temp_data <- rnorm(n = 50, mean = 15, sd = 3)
temp_data
```

```{r}
length(temp_data)
```

```{r}
mean(temp_data)
```

```{r}
sd(temp_data)
```

```{r}
class(temp_data)
```

```{r}
names <- c("Fluffy", "Jesse")
names <- c('Fluffy', 'Jesse')
class(names)
```

## Factor data

```{r}
# Generate a vector of vegetation types

vegetation <- sample(c("grassland", "forest", "wetland"), size = 50, replace = TRUE)

# View the vegetation data
vegetation
```

```{r}
class(vegetation)
```

```{r}
veg_fact <- as.factor(vegetation)
class(veg_fact)
```

## Ordered factors

```{r}
# Vector of ordinal data representing the successional stage of a forest
coffee <- c("Small", "Tall", "Venti", "Grande", "Ginormous")
class(coffee)
```

```{r}
# Convert to ordered factor
coffee <- factor(coffee, ordered = TRUE,
                 levels = c("Small", "Tall", "Venti",
                            "Grande", "Ginormous"))
coffee
```

```{r}
class(coffee)
```

## Logical data

```{r}
# Generate a vector of 1s and 0s to represent the presence
# or absence of a species in different ecological sites
species_presence <- sample(c(0, 1), 10,
                           replace = TRUE)
class(species_presence)
```

```{r}
species_presence_logi <- as.logical(species_presence)
class(species_presence_logi)
```

## Dates


```{r}
my_time <- as.POSIXct("2022-03-10 12:34:56")
class(my_time)
```

## Missing values

```{r}
# Set the length of the sequence
n <- 100

# Generate a sequence of random normal numbers with
# mean 0 and standard deviation 1
data <- rnorm(n, mean = 0, sd = 1)

# Randomly assign 5% of the values as missing
missing_indices <- sample(1:n, size = round(0.05*n))
data[missing_indices] <- NA
length(data)
```

```{r}
mean(data, na.rm = TRUE)
```

```{r}
length(data)
data_sans_na <- na.omit(data)
length(data_sans_na)
```

## Matrices (plural) and a matric (singular)

```{r}
# create a numeric matrix
my_matrix_1 <- matrix(1:6, nrow = 2, ncol = 3,
                    byrow = FALSE)

my_matrix_2 <- matrix(1:6, nrow = 2, ncol = 3,
                    byrow = TRUE)
```

```{r}
# print the matrix
my_matrix_1
my_matrix_2
```

```{r}
dim(my_matrix_1)
```

```{r}
mat <- matrix(rnorm(n = 15, mean = 0, sd = 1),
              nrow = 5, ncol = 3)
mat <- round(mat, 2)
dim(mat)
ncol(mat)
nrow(mat)
```

```{r}
# coerce the matrixto a vector
# 2D to 1D
vect <- as.vector(mat)
vect
length(vect)
```

```{r}
mat
```

```{r}
mat[2, ]
mat[5, c(1, 3)]
```

## Arrays

```{r}
# create a 2-dimensional array
my_array <- array(1:27, dim = c(3, 3, 3))

# print the array
my_array
```

```{r}
dim(my_array)
```

```{r}
my_array[2, 2, 2]

# square bracket notation:
# [row, column, depth]
my_array[2, 3, 3]
```

```{r}
as.vector(my_array)
```

## Making a data.frame

```{r}
# Create a vector of dates
dates <- as.Date(c("2022-01-01", "2022-01-02", "2022-01-03",
                   "2022-01-04", "2022-01-05"))

# Create a vector of numeric data
numeric_data <- rnorm(n = 5, mean = 0, sd = 1)

# Create a vector of categorical data
categorical_data <- c("A", "B", "C", "A", "B")

# Combine the vectors into a data.frame
my_dataframe <- data.frame(day = dates,
                           number = numeric_data,
                           category = categorical_data)

# Print the dataframe
my_dataframe
```

```{r}
class(my_dataframe)
colnames(my_dataframe)
rownames(my_dataframe)
dim(my_dataframe)
str(my_dataframe)
mean(my_dataframe$number)
range(my_dataframe$day)
summary(my_dataframe)
glimpse(my_dataframe)
```

```{r}
my_tibble <- as_tibble(my_dataframe)
class(my_tibble)
my_tibble
```

## Exercise in frustration

```{r}
read_csv("../data/SAMOS/other_data/SACTN_SAWS.csv") |> 
  mutate(month = as.integer(month(date, label = FALSE))) |> 
  group_by(site) |> 
  mutate(temp_y = mean(temp, na.rm = TRUE)) |> 
  group_by(site, month) |> 
  summarise(temp_m = mean(temp, na.rm = TRUE),
            temp_y = mean(temp_y)) |> 
  ggplot(aes(x = month)) +
    geom_line(aes(y = temp_y), colour = "red") +
    geom_line(aes(y = temp_m)) +
    scale_x_continuous(breaks = c(3, 6, 9, 12)) +
    facet_wrap(~site)


```

## Reading in data into R

For this exercise, we will use the `laminaria.csv` file. It is located in the `data/SAMOS/other_data` folder.

```{r}
laminaria <- read_csv("../data/SAMOS/other_data/laminaria.csv")
```

```{r}
laminaria
head(laminaria, 7)
tail(laminaria, 2)
summary(laminaria)
glimpse(laminaria)
names(laminaria)
```

```{r}
library(skimr)
skim(laminaria) # using built-in `iris` data
```

## Tidyverse sneak peak

```{r}
# Note: output NOT assigned to an object (no `<-`):
laminaria %>% # Tell R which dataframe you are using ("%>%" == "|>" )
  select(site, total_length) # Select only specific columns
```

```{r}
laminaria %>% 
  select(site, total_length) %>% # Select specific columns first
  slice(56:78)
# what does the '56:78' do? Change some numbers and run the code again. What happens?
```

```{r}
laminaria %>%
  filter(site == "Kommetjie")
```

```{r}
laminaria |> 
  filter(blade_weight >= 1.5)
```

```{r}
lam_kom <- laminaria %>%
  filter(site == "Kommetjie")
```

```{r}
laminaria %>% # Tell R which dataset to use
  filter(site == "Kommetjie") %>% # Filter out only records from Kommetjie
  nrow() # Count the number of remaining rows
```

```{r}
laminaria %>% # Tell R which dataset to use
  filter(total_length == max(total_length)) # Select row with max total length
```

## Some new functions: summarise, mutate, and group_by

```{r}
laminaria %>% # Chose the dataframe
  summarise(avg_bld_len = mean(blade_length)) # Calculate mean blade length
```

```{r}
laminaria %>%
  summarise(avg_stp_ln = round(mean(total_length), 1),
            sd_stp_ln = round(sd(total_length), 1))
```

```{r}
laminaria %>% 
  summarise(avg_stp_ms = mean(stipe_mass, na.rm = TRUE))
```

```{r}
res <- laminaria %>%
  group_by(site) %>%
  summarise(var_bl = var(blade_length),
            n_bl = n(),
            sd_bl = sd(blade_length)) %>%
  mutate(se_bl = var_bl / sqrt(n_bl))
```

```{r}
res
```

```{r}
write.csv(res, "../data/SAMOS/other_data/laminaria_results.csv")
```

## Graphics with R

```{r}
data()
```

```{r}
# Load data
ChickWeight <- datasets::ChickWeight

# Create a basic figure
ggplot(data = ChickWeight, aes(x = Time, y = weight)) +
  geom_line(aes(group = Chick)) +
  geom_point(colour = "red")
```

```{r}
ggplot(data = ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
  geom_point() +
  geom_line(aes(group = Chick))
```

```{r}
ggplot(data = ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
  geom_point() +
  geom_smooth(method = "gam")
```

```{r}
ggplot(data = ChickWeight, aes(x = Time, y = weight)) +
  geom_point(colour = "blue") +
  geom_line(aes(group = Chick), colour = "blue")
```

```{r}
ggplot(data = ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
  geom_point(aes(size = weight)) +
  geom_smooth(method = "lm", linewidth = 2.0)
```

```{r}
ggplot(data = ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Days", y = "Mass (g)", colour = "diet type") + # Change the labels
  theme(legend.position = "bottom") # Change the legend position
```

